---
title: "TIP_Equity24"
author: "Margaret Atkinson"
output: html_document
updated_date: "2024-01-23"
updated_by: "Seth Strumwasser"
---

# TIP Demographics
This script takes as input a table of area fractions by GEOID representing the proportion of that GEOID
we want to count for each TIP project. In the script we follow these steps:

1. Pull census data
2. Apply area fractions
3. Summarize by project
4. Combine tables
5. Arrange output table


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

2. Libraries
```{r}
library(dplyr)
library(tidycensus)
library(tidyverse)
library(stringr)
library(ggplot2)
library(readr)
```

```{r}
# ****** configuration - check all these variables before running!!*********

# input_blkgrps_table <- read_csv("\\Users\\sstrumwasser.AD\\Documents\\ArcGIS\\Projects\\needs_assessment_proximity\\arcpy_outputs\\ebikes_bg_af.csv")
# input_tracts_table <- read_csv("\\Users\\sstrumwasser.AD\\Documents\\ArcGIS\\Projects\\needs_assessment_proximity\\arcpy_outputs\\ebikes_tract_af.csv")

# if you have multiple input geometries, combine them here:
#
# input_blkgrps_table <- do.call("rbind", list( - BLOCK GROUP TABLES HERE - ))
# input_tracts_table <- do.call("rbind", list( - TRACT TABLES HERE - ))

results_dir <- "\\Users\\sstrumwasser.AD\\Documents\\ArcGIS\\Projects\\TIP_25_29\\results\\"

input_blkgrps_table <- do.call("rbind", list(
  read_csv(paste0(results_dir, "\\lines_af_bg_eighth_mi.csv")),
  read_csv(paste0(results_dir, "\\lines_af_bg_quart_mi.csv")),
  read_csv(paste0(results_dir, "\\lines_af_bg_half_mi.csv")),
  read_csv(paste0(results_dir, "\\polygons_af_bg.csv"))
))

input_tracts_table <- do.call("rbind", list(
  read_csv(paste0(results_dir, "\\lines_af_tract_eighth_mi.csv")),
  read_csv(paste0(results_dir, "\\lines_af_tract_quart_mi.csv")),
  read_csv(paste0(results_dir, "\\lines_af_tract_half_mi.csv")),
  read_csv(paste0(results_dir, "\\polygons_af_tract.csv"))
))

proj_id_field_name <- "CTPS_ID_text"

output_table <- "\\Users\\sstrumwasser.AD\\Documents\\ArcGIS\\Projects\\TIP_25_29\\results\\demographics_output.csv"
```


3. Standard Error Function
```{r}
# standard error proportion fxn
std_err_proportion <- function(output, numerator, denominator, std_err_numerator, std_err_denominator){
  numerator <- as.numeric(numerator)
  denominator <- as.numeric(denominator)
  std_err_denominator <- as.numeric(std_err_denominator)
  std_err_numerator <- as.numeric(std_err_numerator)
  if(denominator == 0){
    result <- 0
  }else{
    radicand <- (std_err_numerator^2) - ((numerator / denominator)^2)  *  (std_err_denominator^2)
    if(radicand < 0){
      radicand <- (std_err_numerator^2) + ((numerator / denominator)^2)  *  (std_err_denominator^2)
    }
    if(numerator == denominator){
      result <- round(std_err_numerator / denominator, 5)
    } else if(radicand == 0) {
      result <- 0
    } else{
      result <- round((radicand ^0.5) / denominator, 5)
    }
    
  }
}
#(written by Paul Reim)
```
4. Census API key set up
```{r}
# Census API key set up
#census_api_key("e4bec76221ba04c7df76c7c580659bf1f54ed2c1", install = TRUE)
# First time, reload your environment so you can use the key without restarting R.
readRenviron("~/.Renviron")
# You can check it with:
Sys.getenv("CENSUS_API_KEY")
```
Prepare ACS 2020 Field Categories
```{r}
# ACS 2022 Fields
# universe
acs_noninst_pop <- c("b18101_001e")

# senior
acs_sr_noninst_pop <- c("b18101_015e","b18101_018e","b18101_034e","b18101_037e")

# disabled
acs_disabled_pop <- c("b18101_004e","b18101_007e","b18101_010e","b18101_013e","b18101_016e","b18101_019e",
                      "b18101_023e","b18101_026e","b18101_029e","b18101_032e","b18101_035e","b18101_038e")

# senior disabled
acs_sr_disab_pop <- c("b18101_016e","b18101_019e","b18101_035e","b18101_038e")

# lep universe
acs_pop_5p <- c("b16004_001e")

# lep populations
acs_lep_pop <- c("b16004_001e","b16004_003e","b16004_005e","b16004_010e","b16004_015e","b16004_020e","b16004_025e","b16004_027e","b16004_032e","b16004_037e",
                 "b16004_042e","b16004_047e","b16004_049e","b16004_054e","b16004_059e","b16004_064e")

# poverty status determined estimate
pov_stat_det_est <- c("c17002_001e")
# poverty status determined margin of error
pov_stat_det_moe <- c("c17002_001m")

# income less than 2x poverty level population
lt_2x_pov_est <- c("c17002_001e","c17002_008e")
lt_2x_pov_moe <- c("c17002_001m","c17002_008m")

# youth and older adults
acs_pop75p <- c("B01001_023E","B01001_024E","B01001_025E","B01001_047E","B01001_048E","B01001_049E")
acs_pop18u <- c("B01001_001E","B01001_003E","B01001_004E","B01001_005E","B01001_006E","B01001_027E","B01001_028E","B01001_029E","B01001_030E")


#all acs fields
acs_fields <- c(acs_noninst_pop,acs_sr_noninst_pop,acs_disabled_pop,
               acs_pop_5p,acs_lep_pop,pov_stat_det_est,
               pov_stat_det_moe,lt_2x_pov_est,lt_2x_pov_moe,
               acs_pop75p,acs_pop18u)
print("Done.")  
```

Download tables from tidycensus
```{r}
# Grab Census 2020 geography, total population, minority data, older adults, youth
# download block groups and minority data and total population data
decennial_yr <- 2020

bg20 <- 
  get_decennial(geography = 'block group', 
                variables = c('P5_001N', 'P5_003N', # white and minority
                              'P12_001N', 'P12_023N', 'P12_024N', 'P12_025N', 'P12_047N', 
                              'P12_048N', 'P12_049N', # Older adults
                              'P12_003N', 'P12_004N', 'P12_005N', 'P12_006N', 'P12_027N',  
                              'P12_028N', 'P12_029N', 'P12_030N'), # Youth 
                year = decennial_yr, 
                sumfile = "dhc", 
                state = "MA", 
                output = 'wide')
```
ACS
```{r}
acs_yr <- 2022

# download block group geography and demographic data
acsbgs <- get_acs(geography = 'block group', 
                  variables = toupper(acs_fields), 
                  year = acs_yr, 
                  state = "MA", 
                  output = 'wide')
# download tract geography and demographic data
acstrts <- get_acs(geography = 'tract', 
                   variables = toupper(acs_fields), 
                   year = acs_yr, 
                   state = "MA", 
                   output = 'wide')

```

Calculate counts for Census - total pop and minority pop 

```{r}
#total population
bg20$TotalPop <- bg20$P5_001N

#minority population
bg20$MinPop <- bg20$TotalPop - bg20$P5_003N

# Older adults and youth
#
# Note: P12_001N and P5_001N are equivalent
bg20 <- bg20 %>% 
  rowwise() %>% 
  mutate(OldPop = sum(P12_023N, P12_024N, P12_025N, P12_047N, 
                      P12_048N, P12_049N),
         YouthPop = sum(P12_003N, P12_004N, P12_005N, P12_006N, P12_027N,  
                              P12_028N, P12_029N, P12_030N))

```

ACS - Block Groups
``` {r, warning = FALSE} 
#Pov_Stat_Det_Pop (Taken from ACS because only using % and not released from 2020 Census yet)
####POV STAT DET & LT 2X POV####
acsbgs$pov_stat_det_est <- acsbgs$C17002_001E
acsbgs$pov_stat_det_moe <- acsbgs$C17002_001M

acsbgs$lt_2x_pov_est <- acsbgs$C17002_001E - acsbgs$C17002_008E

#std_err_sum (e.g. standard error of a sum or difference)
# std_err_sum =  round(((stderr1^2)+(stderr2^2)) * 0.5,0)
acsbgs$lt_2x_pov_moe <- round(((acsbgs$C17002_001M^2)+(acsbgs$C17002_008M^2)) * 0.5,0)

acsbgs <- acsbgs %>% mutate(lt_2x_pov_shr_est = ifelse(pov_stat_det_est == 0, 0, round(lt_2x_pov_est/pov_stat_det_est, 5)))

#std_err_proportion (e.g. standard error of a proportion)
acsbgs$lt_2x_pov_shr_moe <- apply(
   acsbgs[,c('lt_2x_pov_est','pov_stat_det_est', 'lt_2x_pov_moe', 'pov_stat_det_moe')], 
   1, 
   function(x) std_err_proportion(
    numerator=x[1],
    denominator=x[2],
    std_err_numerator=x[3],
    std_err_denominator = x[4]
    )
   )



####LEP####
acsbgs$acs_pop_5p <- acsbgs$B16004_001E

acsbgs$acs_lep_pop <- (acsbgs$B16004_001E - acsbgs$B16004_003E - acsbgs$B16004_005E - acsbgs$B16004_010E - acsbgs$B16004_015E - acsbgs$B16004_020E - acsbgs$B16004_025E - acsbgs$B16004_027E - acsbgs$B16004_032E - acsbgs$B16004_037E - acsbgs$B16004_042E - acsbgs$B16004_047E - acsbgs$B16004_049E - acsbgs$B16004_054E - acsbgs$B16004_059E - acsbgs$B16004_064E) 

acsbgs$bg_totpop <- acsbgs$B01001_001E

```



ACS - Tracts
```{r}
####DISABLED POP ACS####
acstrts$acs_disab_pop <- (acstrts$B18101_004E + acstrts$B18101_007E + acstrts$B18101_010E +  
                          acstrts$B18101_013E + acstrts$B18101_016E + acstrts$B18101_019E +
                          acstrts$B18101_023E + acstrts$B18101_026E + acstrts$B18101_029E +
                          acstrts$B18101_032E + acstrts$B18101_035E + acstrts$B18101_038E)

acstrts$acs_noninst_pop <- acstrts$B18101_001E

```

Import the TIP data
```{r}
# tables for each input geometry/census geometry combination
bg_cen20 <- input_blkgrps_table

tract_cen20 <- input_tracts_table
```

Join TIP data to demographic data (Block Groups and Tracts)
```{r}

bg_acs <- merge(bg_cen20 %>% 
                  select(geoid, area_fraction, all_of(proj_id_field_name)), acsbgs, 
                  by.x =char('geoid'), by.y ='GEOID')
bg_acs[is.na(bg_acs)] <- 0

tract_acs <- merge(tract_cen20 %>% 
                     select(geoid, area_fraction, all_of(proj_id_field_name)), acstrts, 
                     by.x ='geoid', by.y ='GEOID')
tract_acs[is.na(tract_acs)] <- 0

bg_cen <- merge(bg_cen20 %>% 
                  select(geoid, area_fraction, all_of(proj_id_field_name)), bg20, by.x ='geoid', 
                  by.y ='GEOID')
bg_cen[is.na(bg_cen)] <- 0

```


Apply Area Fraction to BGs and Tracts 
```{r}
#2020 Census
bg_cen$TotalPop_AF <- round(bg_cen$P5_001N  *  bg_cen$area_fraction)
bg_cen$MinPop_AF   <- round(bg_cen$MinPop  *  bg_cen$area_fraction)
bg_cen$OldPop_AF   <- round(bg_cen$OldPop  *  bg_cen$area_fraction)
bg_cen$YouthPop_AF   <- round(bg_cen$YouthPop  *  bg_cen$area_fraction)

#2022 ACS - BGs
#POV & 2x POV level
bg_acs$pov_stat_det_est_AF <- bg_acs$pov_stat_det_est  *  bg_acs$area_fraction
bg_acs$lt_2x_pov_est_AF    <- bg_acs$lt_2x_pov_est  *  bg_acs$area_fraction
bg_acs$lt_2x_pov_moe_AF    <- bg_acs$lt_2x_pov_moe  *  bg_acs$area_fraction


####LEP####
bg_acs$acs_pop_5p_AF  <- bg_acs$acs_pop_5p * bg_acs$area_fraction
bg_acs$acs_lep_pop_AF <- bg_acs$acs_lep_pop * bg_acs$area_fraction

####ACS AGE AND HH#####
bg_acs$bg_totpop_AF <- bg_acs$bg_totpop * bg_acs$area_fraction


#2022 ACS - Tracts
tract_acs$acs_disab_pop_AF <- tract_acs$acs_disab_pop * tract_acs$area_fraction
tract_acs$acs_noninst_pop_AF <- tract_acs$acs_noninst_pop * tract_acs$area_fraction

```


Sum by Project
```{r}
buf_bg_cen <- bg_cen %>% 
  group_by_at(proj_id_field_name) %>% 
  summarise(
    TotalPop = sum(TotalPop_AF),
    MinPop = sum(MinPop_AF),
    OldPop = sum(OldPop_AF),
    YouthPop = sum(YouthPop_AF)
  )


buf_bg_acs <- bg_acs %>% group_by_at(proj_id_field_name) %>% summarise(
  pov_stat_det_est = sum(pov_stat_det_est_AF),
  lt_2x_pov_est = sum(lt_2x_pov_est_AF),
  lt_2x_pov_moe = sum(lt_2x_pov_moe_AF),
  acs_pop_5p = sum(acs_pop_5p_AF),
  acs_lep_pop = sum(acs_lep_pop_AF),
  bg_totpop = sum(bg_totpop_AF)
)


buf_tract_acs <- tract_acs %>% 
  group_by_at(proj_id_field_name) %>% 
  summarise(acs_disab_pop = sum(acs_disab_pop_AF),
            acs_noninst_pop = sum(acs_noninst_pop_AF))

```


Calculate Percentages
```{r}

buf_bg_cen$MinPerc <- round(buf_bg_cen$MinPop/buf_bg_cen$TotalPop,2)
buf_bg_cen$OldPerc <- round(buf_bg_cen$OldPop/buf_bg_cen$TotalPop,2)
buf_bg_cen$YouthPerc <- round(buf_bg_cen$YouthPop/buf_bg_cen$TotalPop,2)

buf_bg_acs$lt_2x_pov_perc <- round(buf_bg_acs$lt_2x_pov_est / buf_bg_acs$pov_stat_det_est,2)

buf_bg_acs$lep_perc <- round(buf_bg_acs$acs_lep_pop / buf_bg_acs$acs_pop_5p,2)

buf_tract_acs$disab_perc <- round(buf_tract_acs$acs_disab_pop / buf_tract_acs$acs_noninst_pop,2)

```

Import MPO totals
```{r}
mpo_totals <- read_csv("\\Users\\sstrumwasser.AD\\Documents\\GitHub\\TIP_Demographics\\mpo_pops.csv")
```


Export file
```{r}
# merge all block group tables
buf_bg <- merge(buf_bg_acs, buf_bg_cen, by=proj_id_field_name)

# merge block group tables with tract table
all_vars <- merge(buf_bg, buf_tract_acs, by=proj_id_field_name)

# round values
all_vars_round_pops <- all_vars %>%
  mutate(
    MinPop_round = round(MinPop,0),
    TotalPop_round = round(TotalPop,0),
    OldPop_round = round(OldPop, 0),
    YouthPop_round = round(YouthPop, 0),
    lt_2x_pov_est_round = round(lt_2x_pov_est,0),
    lep_round = round(acs_lep_pop),
    disab_round = round(acs_disab_pop,0),
    .keep = c("all")
  )

# possible_fields <- c(TotalPop_round,
#   MinPop_round,
#   MinPerc,
#   lt_2x_pov_est_round,
#   lt_2x_pov_perc,
#   lep_round,
#   lep_perc,
#   u18_round,
#   u18_perc,
#   p75_round,
#   p75_perc,
#   disab_round,
#   disab_perc)


final <- all_vars_round_pops %>% select(
  all_of(proj_id_field_name),
  TotalPop_round,
  MinPop_round,
  MinPerc,
  OldPop_round,
  OldPerc,
  YouthPop_round,
  YouthPerc,
  lt_2x_pov_est_round,
  lt_2x_pov_perc,
  lep_round,
  lep_perc,
  disab_round,
  disab_perc
  )

write.csv(final, output_table)

```

```{r}
# qc
all_bgs <- read.csv(input_blkgrps_table)
bg_cen_qc <- merge(all_bgs %>% select(geoid), bg20, by.x ='geoid', by.y ='GEOID')
sum_test <- sum(bg_cen_qc$P5_001N)
# write.csv(bg_cen_qc, "\\Users\\sstrumwasser.AD\\Documents\\ArcGIS\\Projects\\needs_assessment_proximity\\R_outputs\\charging_stations_half_mile_equity_qc.csv")
```

State totals for demographics project
```{r}
cen_state <- bg20 %>% 
  ungroup() %>% 
  summarize(
    TotalPop = sum(TotalPop),
    MinPop = sum(MinPop),
    YouthPop = sum(YouthPop),
    OldPop = sum(OldPop)
    )

acs_bg_state <- acsbgs %>% 
  ungroup() %>% 
  summarize(
    low_inc = sum(lt_2x_pov_est),
    acs_lep_pop = sum(acs_lep_pop)
  )

acs_tract_state <- acstrts %>% 
  ungroup() %>% 
  summarize(
    acs_disab_pop = sum(acs_disab_pop)
  )

state_pops <- cbind(cen_state, acs_bg_state, acs_tract_state)
write_csv(state_pops, "\\Users\\sstrumwasser.AD\\Documents\\ArcGIS\\Projects\\TIP_25_29\\results\\statewide_demographics.csv")
```



