---
title: "TIP_Equity23"
author: "Margaret Atkinson"
output: html_document
updated_date: "2022-11-16"
updated_by: "Seth Strumwasser"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

2. Libraries
```{r}
library(dplyr)
library(tidycensus)
library(tidyverse)
library(stringr)
library(ggplot2)
library(readr)
```
3. Standard Error Function
```{r}
# standard error proportion fxn
std_err_proportion <- function(output, numerator, denominator, std_err_numerator, std_err_denominator){
  numerator <- as.numeric(numerator)
  denominator <- as.numeric(denominator)
  std_err_denominator <- as.numeric(std_err_denominator)
  std_err_numerator <- as.numeric(std_err_numerator)
  if(denominator == 0){
    result <- 0
  }else{
    radicand <- (std_err_numerator^2) - ((numerator / denominator)^2) * (std_err_denominator^2)
    if(radicand < 0){
      radicand <- (std_err_numerator^2) + ((numerator / denominator)^2) * (std_err_denominator^2)
    }
    if(numerator == denominator){
      result <- round(std_err_numerator / denominator, 5)
    } else if(radicand == 0) {
      result <- 0
    } else{
      result <- round((radicand ^0.5) / denominator, 5)
    }
    
  }
}
#(written by Paul Reim)
```
4. Census API key set up
```{r}
# Census API key set up
#census_api_key("e4bec76221ba04c7df76c7c580659bf1f54ed2c1", install = TRUE)
# First time, reload your environment so you can use the key without restarting R.
readRenviron("~/.Renviron")
# You can check it with:
# Sys.getenv("CENSUS_API_KEY")
```
Prepare ACS 2020 Field Categories
```{r}
# ACS 2020 Fields
# universe
acs_noninst_pop <- c("b18101_001e")

# senior
acs_sr_noninst_pop <- c("b18101_015e","b18101_018e","b18101_034e","b18101_037e")

# disabled
acs_disabled_pop <- c("b18101_004e","b18101_007e","b18101_010e","b18101_013e","b18101_016e","b18101_019e",
                      "b18101_023e","b18101_026e","b18101_029e","b18101_032e","b18101_035e","b18101_038e")

# senior disabled
acs_sr_disab_pop <- c("b18101_016e","b18101_019e","b18101_035e","b18101_038e")

# number of households
acs_hh <- c("b25044_001e")

# 0 vehicle households
acs_0veh_hh <- c("b25044_003e","b25044_010e")

# low income households
acs_low_inc_hh <- c("b19001_001e","b19001_002e","b19001_003e","b19001_004e","b19001_005e","b19001_006e","b19001_007e","b19001_008e","b19001_009e","b19001_010e")

# lep universe
acs_pop_5p <- c("b16004_001e")

# lep populations
acs_lep_pop <- c("b16004_001e","b16004_003e","b16004_005e","b16004_010e","b16004_015e","b16004_020e","b16004_025e","b16004_027e","b16004_032e","b16004_037e",
                 "b16004_042e","b16004_047e","b16004_049e","b16004_054e","b16004_059e","b16004_064e")

# poverty status determined estimate
pov_stat_det_est <- c("c17002_001e")
# poverty status determined margin of error
pov_stat_det_moe <- c("c17002_001m")

# income less than 2x poverty level population
lt_2x_pov_est <- c("c17002_001e","c17002_008e")
lt_2x_pov_moe <- c("c17002_001m","c17002_008m")

# 16 or older population
acs_pop16p <- c("b23001_001e")


acs_workers16p <- c("b08301_001e")  
acs_workers16p_shr <- c("b23001_001e","b08301_001e")
acs_pop75p <- c("B01001_023E","B01001_024E","B01001_025E","B01001_047E","B01001_048E","B01001_049E")
acs_pop18u <- c("B01001_001E","B01001_003E","B01001_004E","B01001_005E","B01001_006E","B01001_027E","B01001_028E","B01001_029E","B01001_030E")


#all acs fields
acs_fields <- c(acs_noninst_pop,acs_sr_noninst_pop,acs_disabled_pop,acs_sr_disab_pop,
               acs_hh,acs_0veh_hh,acs_low_inc_hh,acs_pop_5p,acs_lep_pop,pov_stat_det_est,
               pov_stat_det_moe,lt_2x_pov_est,lt_2x_pov_moe,acs_pop16p,acs_workers16p,
               acs_workers16p_shr,acs_pop75p,acs_pop18u)
print("Done.")  
```

Download tables from tidycensus
```{r}
# Grab Census 2020 geography, total population, and minority data
# download block groups and minority data and total population data
bg20 <- get_decennial(geography = 'block group', variables = c('P1_001N','P1_003N','P2_005N'), year = 2020, sumfile = "pl", state = "MA", output = 'wide')
```
ACS
```{r}
acs_yr <- 2020

# download block group geography and demographic data
acsbgs <- get_acs(geography = 'block group', variables = toupper(acs_fields), year = acs_yr, state = "MA", output = 'wide')
# download tract geography and demographic data
acstrts <- get_acs(geography = 'tract', variables = toupper(acs_fields), year = acs_yr, state = "MA", output = 'wide')

```

Calculate counts for Census - total pop and minority pop 

```{r}
#total population
bg20$TotalPop <- bg20$P1_001N
#minority population
bg20$MinPop <- bg20$TotalPop - bg20$P2_005N

```

ACS - Block Groups
``` {r, warning = FALSE} 
#Pov_Stat_Det_Pop (Taken from ACS because only using % and not released from 2020 Census yet)
####POV STAT DET & LT 2X POV####
acsbgs$pov_stat_det_est <- acsbgs$C17002_001E
acsbgs$pov_stat_det_moe <- acsbgs$C17002_001M

acsbgs$lt_2x_pov_est <- acsbgs$C17002_001E - acsbgs$C17002_008E

#std_err_sum (e.g. standard error of a sum or difference)
# std_err_sum =  round(((stderr1^2)+(stderr2^2))*0.5,0)
acsbgs$lt_2x_pov_moe <- round(((acsbgs$C17002_001M^2)+(acsbgs$C17002_008M^2))*0.5,0)

acsbgs <- acsbgs %>% mutate(lt_2x_pov_shr_est = ifelse(pov_stat_det_est == 0, 0, round(lt_2x_pov_est/pov_stat_det_est, 5)))

#std_err_proportion (e.g. standard error of a proportion)
acsbgs$lt_2x_pov_shr_moe <- apply(
   acsbgs[,c('lt_2x_pov_est','pov_stat_det_est', 'lt_2x_pov_moe', 'pov_stat_det_moe')], 
   1, 
   function(x) std_err_proportion(
    numerator=x[1],
    denominator=x[2],
    std_err_numerator=x[3],
    std_err_denominator = x[4]
    )
   )

####HOUSEHOLDS - 0VEH AND LOWINC####
acsbgs$acs_hh <- acsbgs$B25044_001E

acsbgs$acs_0veh_hh <- acsbgs$B25044_003E + acsbgs$B25044_010E

acsbgs$acs_0veh_hh_shr <- round(acsbgs$acs_0veh_hh/acsbgs$acs_hh,5)

acsbgs$acs_tot_hh <- acsbgs$B19001_001E

# Note: 392/5000 is most likely the share of the households in that last category
# that are low-income. This would be calculated elsewhere and reported here.
# TODO: Check whether we can calculate the 392/5000 at runtime.
acsbgs$acs_low_inc_hh <- round(acsbgs$B19001_002E + acsbgs$B19001_003E + acsbgs$B19001_004E + acsbgs$B19001_005E +
                               acsbgs$B19001_006E + acsbgs$B19001_007E + acsbgs$B19001_008E + acsbgs$B19001_009E +
                               (392/5000)*acsbgs$B19001_010E,0)


####LEP####
acsbgs$acs_pop_5p <- acsbgs$B16004_001E

acsbgs$acs_lep_pop <- (acsbgs$B16004_001E-acsbgs$B16004_003E-acsbgs$B16004_005E-acsbgs$B16004_010E-acsbgs$B16004_015E-
                       acsbgs$B16004_020E-acsbgs$B16004_025E-acsbgs$B16004_027E-acsbgs$B16004_032E-acsbgs$B16004_037E-
                       acsbgs$B16004_042E-acsbgs$B16004_047E-acsbgs$B16004_049E-acsbgs$B16004_054E-acsbgs$B16004_059E-
                       acsbgs$B16004_064E) 



####ACS AGE AND HH#####

acsbgs$bg_pop75p <- (acsbgs$B01001_023E+acsbgs$B01001_024E+acsbgs$B01001_025E+acsbgs$B01001_047E+acsbgs$B01001_048E+acsbgs$B01001_049E)

acsbgs$bg_pop18u <- (acsbgs$B01001_003E+acsbgs$B01001_004E+acsbgs$B01001_005E+acsbgs$B01001_006E+
                                          acsbgs$B01001_027E+acsbgs$B01001_028E+acsbgs$B01001_029E+acsbgs$B01001_030E)

acsbgs$bg_totpop <- acsbgs$B01001_001E




```



ACS - Tracts
```{r}
####DISABLED POP ACS####
acstrts$acs_noninst_pop <- acstrts$B18101_001E

acstrts$acs_sr_noninst_pop <- (acstrts$B18101_015E +acstrts$B18101_018E +acstrts$B18101_034E +acstrts$B18101_037E)

acstrts$acs_disab_pop <- (acstrts$B18101_004E + acstrts$B18101_007E + acstrts$B18101_010E + acstrts$B18101_013E +
                         acstrts$B18101_016E + acstrts$B18101_019E + acstrts$B18101_023E + acstrts$B18101_026E +
                         acstrts$B18101_029E + acstrts$B18101_032E + acstrts$B18101_035E + acstrts$B18101_038E)

acstrts$acs_sr_disab_pop <- (acstrts$B18101_016E + acstrts$B18101_019E + acstrts$B18101_035E + acstrts$B18101_038E)


acstrts$acs_pop16p <- acstrts$B23001_001E

acstrts$acs_workers16p <- acstrts$B08301_001E

```

Import the TIP data
```{r}
#tip projects with three different geography types
# tip_bg_acs19 <- read_csv("G:/Data_Resources/tip stuff/TIP_scoring/ffy2023/tip_bg_acs19.csv")
# tip_bg_cen20 <- read_csv("G:/Data_Resources/tip stuff/TIP_scoring/ffy2023/tip_bg_cen20.csv")
# tip_tract_acs19 <- read_csv("G:/Data_Resources/tip stuff/TIP_scoring/ffy2023/tip_tract_acs19.csv")

tip_bg_cen20 <- read.csv("C:/Users/sstrumwasser.AD/Documents/TIP_Equity23/arcpy_outputs/TIP_projects_lines_netBufPoly_blkgrps_111422.csv")

tip_tract_cen20 <- read.csv("C:/Users/sstrumwasser.AD/Documents/TIP_Equity23/arcpy_outputs/TIP_projects_lines_tracts_test_111422.csv")

# tip_bg_acs19 <- read_csv("./data/tip_bg_acs19.csv")
# tip_bg_cen20 <- read_csv("./data/tip_bg_cen20.csv")
# tip_tract_acs19 <- read_csv("./data/tip_tract_acs19.csv")


# #rebuild the GEOIDs to match
# tip_bg_acs19 <- tip_bg_acs19 %>% 
#   mutate(
#     geoid_old = geoid, # keep the original one
#     # geoid = paste0(
#     # str_pad(state, 2, pad = 0),
#     # str_pad(county, 3, pad = 0),
#     # str_pad(tract10, 6, pad = 0),
#     # str_pad(block10, 4, pad = 0))
  # )


# clean up!
rm(acs_noninst_pop,acs_sr_noninst_pop,acs_disabled_pop,acs_sr_disab_pop,
               acs_hh,acs_0veh_hh,acs_low_inc_hh,acs_pop_5p,acs_lep_pop,pov_stat_det_est,
               pov_stat_det_moe,lt_2x_pov_est,lt_2x_pov_moe,acs_pop16p,acs_workers16p,
               acs_workers16p_shr,acs_pop75p,acs_pop18u)

```

Join TIP data to demographic data (Block Groups and Tracts)
```{r}

tip_bg_acs <- merge(tip_bg_cen20 %>% select(geoid, area_fraction,TIP_ID), acsbgs, by.x ='geoid', by.y ='GEOID')
tip_bg_acs[is.na(tip_bg_acs)] <- 0

tip_tract_acs <- merge(tip_tract_cen20 %>% select(geoid, area_fraction,TIP_ID),acstrts, by.x ='geoid', by.y ='GEOID')
tip_tract_acs[is.na(tip_tract_acs)] <- 0

tip_bg_cen <- merge(tip_bg_cen20 %>% select(geoid, area_fraction,TIP_ID), bg20, by.x ='geoid', by.y ='GEOID')
tip_bg_cen[is.na(tip_bg_cen)] <- 0

```


Apply Area Fraction to BGs and Tracts 
```{r}
#2020 Census
tip_bg_cen$TotalPop_AF<- tip_bg_cen$P1_001N*tip_bg_cen$area_fraction
tip_bg_cen$MinPop_AF<- tip_bg_cen$P1_003N*tip_bg_cen$area_fraction

#2020 ACS - BGs
#POV & 2x POV level
tip_bg_acs$pov_stat_det_est_AF <- tip_bg_acs$pov_stat_det_est*tip_bg_acs$area_fraction
tip_bg_acs$lt_2x_pov_est_AF <- tip_bg_acs$lt_2x_pov_est*tip_bg_acs$area_fraction
tip_bg_acs$lt_2x_pov_moe_AF <- tip_bg_acs$lt_2x_pov_moe*tip_bg_acs$area_fraction


####HOUSEHOLDS - 0VEH AND LOWINC####
tip_bg_acs$acs_hh_AF <- tip_bg_acs$acs_hh*tip_bg_acs$area_fraction
tip_bg_acs$acs_0veh_hh_AF <- tip_bg_acs$acs_0veh_hh*tip_bg_acs$area_fraction
tip_bg_acs$acs_0veh_hh_shr_AF <- tip_bg_acs$acs_0veh_hh_shr*tip_bg_acs$area_fraction
tip_bg_acs$acs_tot_hh_AF <- tip_bg_acs$acs_tot_hh*tip_bg_acs$area_fraction
tip_bg_acs$acs_low_inc_hh_AF <- tip_bg_acs$acs_low_inc_hh*tip_bg_acs$area_fraction


####LEP####
tip_bg_acs$acs_pop_5p_AF <- tip_bg_acs$acs_pop_5p*tip_bg_acs$area_fraction
tip_bg_acs$acs_lep_pop_AF <- tip_bg_acs$acs_lep_pop*tip_bg_acs$area_fraction

####ACS AGE AND HH#####
tip_bg_acs$bg_pop75p_AF <- tip_bg_acs$bg_pop75p*tip_bg_acs$area_fraction
tip_bg_acs$bg_pop18u_AF <- tip_bg_acs$bg_pop18u*tip_bg_acs$area_fraction
tip_bg_acs$bg_totpop_AF <- tip_bg_acs$bg_totpop*tip_bg_acs$area_fraction


#2020 ACS - Tracts
tip_tract_acs$acs_noninst_pop_AF <- tip_tract_acs$acs_noninst_pop*tip_tract_acs$area_fraction
tip_tract_acs$acs_sr_noninst_pop_AF <- tip_tract_acs$acs_sr_noninst_pop*tip_tract_acs$area_fraction
tip_tract_acs$acs_disab_pop_AF <- tip_tract_acs$acs_disab_pop*tip_tract_acs$area_fraction
tip_tract_acs$acs_sr_disab_pop_AF <- tip_tract_acs$acs_sr_disab_pop*tip_tract_acs$area_fraction
tip_tract_acs$acs_pop16p_AF <- tip_tract_acs$acs_pop16p*tip_tract_acs$area_fraction
tip_tract_acs$acs_workers16p_AF <- tip_tract_acs$acs_workers16p*tip_tract_acs$area_fraction

```


Sum by Project
```{r}
buf_bg_cen <- tip_bg_cen %>% group_by(TIP_ID) %>% summarise(
  TotalPop = sum(TotalPop_AF),
  MinPop = sum(MinPop_AF)
)


buf_bg_acs <- tip_bg_acs %>% group_by(TIP_ID) %>% summarise(
  pov_stat_det_est = sum(pov_stat_det_est_AF),
  lt_2x_pov_est = sum(lt_2x_pov_est_AF),
  lt_2x_pov_moe = sum(lt_2x_pov_moe_AF),
  acs_hh = sum(acs_hh_AF),
  acs_0veh_hh = sum(acs_0veh_hh_AF),
  acs_0veh_hh_shr = sum(acs_0veh_hh_shr_AF),
  acs_tot_hh = sum(acs_tot_hh_AF),
  acs_low_inc_hh = sum(acs_low_inc_hh_AF),
  acs_pop_5p = sum(acs_pop_5p_AF),
  acs_lep_pop = sum(acs_lep_pop_AF),
  bg_pop75p = sum(bg_pop75p_AF),
  bg_pop18u = sum(bg_pop18u_AF),
  bg_totpop = sum(bg_totpop_AF)
)


buf_tract_acs <- tip_tract_acs %>% 
  group_by(TIP_ID) %>% 
  summarise(
    acs_noninst_pop = sum(acs_noninst_pop_AF),
    acs_sr_noninst_pop = sum(acs_sr_noninst_pop_AF),
    acs_disab_pop = sum(acs_disab_pop_AF),
    acs_sr_disab_pop = sum(acs_sr_disab_pop_AF),
    acs_pop16p = sum(acs_pop16p_AF),
    acs_workers16p = sum(acs_workers16p_AF)
)

```


Calculate Percentages
```{r}

buf_bg_cen$MinPerc <- round(buf_bg_cen$MinPop/buf_bg_cen$TotalPop,2)

buf_bg_acs$lt_2x_pov_perc <- round(buf_bg_acs$lt_2x_pov_est / buf_bg_acs$pov_stat_det_est,2)

buf_bg_acs$hh_0veh_perc <- round(buf_bg_acs$acs_0veh_hh / buf_bg_acs$acs_hh,2)
buf_bg_acs$low_inc_hh_perc <- round(buf_bg_acs$acs_low_inc_hh / buf_bg_acs$acs_tot_hh,2)
buf_bg_acs$lep_perc <- round(buf_bg_acs$acs_lep_pop / buf_bg_acs$acs_pop_5p,2)
buf_bg_acs$p75_perc <- round(buf_bg_acs$bg_pop75p / buf_bg_acs$bg_totpop,2)
buf_bg_acs$u18_perc <- round(buf_bg_acs$bg_pop18u / buf_bg_acs$bg_totpop,2)

buf_tract_acs$disab_perc <- round(buf_tract_acs$acs_disab_pop / buf_tract_acs$acs_noninst_pop,2)
buf_tract_acs$sr_disab_perc <- round(buf_tract_acs$acs_sr_disab_pop / buf_tract_acs$acs_sr_noninst_pop,2)
buf_tract_acs$workers16p_perc <- round(buf_tract_acs$acs_workers16p / buf_tract_acs$acs_pop16p,2)

```

Export file
```{r}
buf_bg <- merge(buf_bg_acs, buf_bg_cen, by="TIP_ID")
all_vars <- merge(buf_bg, buf_tract_acs, by="TIP_ID")
all_vars_round_pops <- all_vars %>%
  mutate(
    MinPop_round = round(MinPop,0),
    TotalPop_round = round(TotalPop,0),
    lt_2x_pov_est_round = round(lt_2x_pov_est,0),
    lep_round = round(acs_lep_pop),
    u18_round = round(bg_pop18u,0),
    p75_round = round(bg_pop75p,0),
    disab_round = round(acs_disab_pop,0),
    .keep = c("all")
  )
final <- all_vars_round_pops %>% select(
  TIP_ID,
  TotalPop_round,
  MinPop_round, 
  MinPerc, 
  lt_2x_pov_est_round,
  lt_2x_pov_perc,
  lep_round,
  lep_perc,
  u18_round,
  u18_perc,
  p75_round,
  p75_perc,
  disab_round,
  disab_perc)

write.csv(final, "C:/Users/sstrumwasser.AD/Documents/TIP_Equity23/tip_demog_pct_2020ACS_2020Cen.csv")

```


